\documentclass[12pt]{article}
\usepackage{amsthm,amssymb,amsmath,mathtools,setspace}
\usepackage{hyperref} % for \texorpdfstring
\onehalfspacing

% --- Module Dependencies ---
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

% --- Load Rosetta Layer (Contracts & Shared Symbols) ---
\input{V_macros.tex}

% Local symbols for this module (index-free)
\providecommand{\Lcal}{\mathcal{L}}
\providecommand{\Scal}{\mathcal{S}}

\title{\textbf{Module 4: The Noether Bridge} \\
\large Validating G4: Derivation of Conserved Currents
\texorpdfstring{\Div{\StressEnergy}=0}{div T = 0}}
\author{G4S — Proof of Conservation (V.tex Compliant)}
\date{\today}

\begin{document}
\maketitle

\section{Input Node: Stationarity from Kinematic Closure (G2)}

The conservation law relies on the \emph{action principle} being stationary. The kinematic
closure \(\GTwoStatement\) guarantees that the fundamental field \(\U\) is a solution to the
Euler–Lagrange equations in the sense needed to invoke stationarity \(\delta \Scal = 0\).

\begin{definition}[Action Functional \(\Scal\)]
Let \(\Lcal\) be a local Lagrangian density encoding informational curvature. The action is
\[
  \Scal[\U] \;=\; \int \Lcal\!\big(\U,\, \grad \U\big)\, d\tau ,
\]
where the measure \(d\tau\) is left abstract (index-free).
\end{definition}

\section{Theorem: Conservation Law (Index-Free Noether)}

\begin{theorem}[Noether Conservation]
For every continuous symmetry \(\Xi\) that leaves the action invariant \((\delta \Scal = 0)\),
there exists an associated current \(\Current\) whose divergence vanishes:
\[
  \Div{\Current} \;=\; 0.
\]
\end{theorem}

\begin{proof}[\textbf{G4 Proof Obligation Fulfillment (Translational Case)}]
\begin{enumerate}
  \item \textbf{Hypothesis (stationarity).} The closure \(\GTwoStatement\) positions \(\U\) at a stationary point: \(\delta \Scal = 0\).
  \item \textbf{Symmetry selection.} Choose translational symmetry \(\Xi_{\mathrm{trans}}\).
  \item \textbf{Noether bridge (API).} Invoke the abstract construction
  \[
    \StressEnergy \;\leftarrow\; \NoetherTensor{\Lcal}{\U}{\Xi_{\mathrm{trans}}}.
  \]
  \item \textbf{Conservation identity.} Symmetry implies the resulting current is conserved:
  \[
    \Div{\StressEnergy} \;=\; 0.
  \]
\end{enumerate}
Thus the Noether bridge carries the geometric constraint \(\GTwoStatement\) into the required
conservation statement, furnishing the bookkeeping law for subsequent thermodynamic closure.
\end{proof}

\end{document}
