\documentclass[12pt,oneside]{book}
\usepackage{amsthm,amssymb,amsmath,mathtools,setspace,thmtools,titling,tocloft}
\usepackage[nogroupskip,nonumberlist]{glossaries}
\makeglossaries

\include{macros}

% Title info
\title{Axioms of Measurement Sufficient for $\Delta S \ge 0$}
\author{Bill Cochran\\wkcochran@gmail.com}
\date{\today}

\begin{document}

\maketitle

\begin{center}
\vspace*{2em}
\begin{center}
\vspace*{2em}
\textit{To Peano, who taught us how to count.}\\[0.4em]
\textit{To Boltzmann, who first counted what could be distinguished.}\\[0.4em]
\textit{To Planck, who taught us that the count is finite.}\\[0.4em]
\textit{To Cantor, who showed us how to count the infinite.}\\[0.4em]
\textit{To Kolmogorov, who showed us that information must be counted to be measured.}\\[0.4em]
\textit{To William of Ockham, who insisted that we only count what is necessary.}
\vspace*{6em}
\begin{quote}
``Je les possède, parce que jamais personne avant moi n’a songé à les posséder.''\\
``Moi, je suis un homme sérieux. Je suis exact. J’aime que l’on soit exact.''\\
\hfill---Antoine de Saint--Exup\'ery, \textit{Le Petit Prince} (1943)
\end{quote}
\end{center}

\vspace*{2em}
\end{center}

\newpage
\frontmatter
\onehalfspacing
\include{front/abstract}
\include{front/toc}
\include{front/roadmap}
\include{front/nb}
\include{front/harm}

\mainmatter
\include{chapters/one}
\include{chapters/two}
\include{chapters/three}
\include{chapters/four}
\include{chapters/five}
\include{chapters/six}
\include{chapters/seven}
\include{chapters/eight}
\backmatter
\include{appendix/proofs}
\include{appendix/notation}

\addcontentsline{toc}{chapter}{Glossary of Definitions and Axioms}
\printglossary

\bibliographystyle{plain}   % or abbrv, alpha, unsrt, etc.
\bibliography{hilbert}   % where references.bib is your BibTeX file
\end{document}
