Trading systems are really an interesting distributed system.  The entire world's
financial marketplace is a loose connection of computer systems operating almost
completely autonomously with no general architect, no central organization, and
highly interconnected.

The use of the Financial Information Exchange protocol has allowed this massive
distributed system to function well enough that bugs and such are not expected
to affect the valuation of nearly everything in the economy.  As such, financial
engineers are required to understand what it means to contribute to the world's
financial markets without disrupting orderly trading.

This book is written to help facilitate everything that is necessary to understand
in order to build a trading system that has minimal likelihood of disrupting orderly
trading while maximizing commercial value. Unfortunately, "disrupting orderly
trading" is insufficient as an engineering specification. This book attempts to tease
out the engineering concerns around maintaining orderly trading while providing
maximal value.

But, beyond the finance, there are engineering tradeoffs as well. This book discusses
the nature of failures in distributed systems and how they manifest in trading. It
also discusses how, and, most importantly, when, to make engineering decisions.
It discusses the major engineering concerns for a variety of common trading systems
and the tradeoffs for each. The patterns expressed may yet still evolve as people
learn how to diversify liquidity.  The goal is to instruct the engineer that 
diversification can be managed through natural organic business growth and to provide
some patterns to apply in selecting technology to meet the requirements.

Of course, alpha decays. Strategies rotate. Success is diversified. Change is the
nature of the business.  There is no single trading system that can do it all.  But,
given enough pieces of a trading system, you can make a trading system that does
neary everything.  And, that is where the flexibility happens.

Decomposition of the trading into systems that can be interconnected and swapped
around provides for very flexible trading environments.  The goal of this book is to
discuss the major responsibilities of a complete trading system, review how these
responsibilities can be encapsulated, 

